manuscript_outline

### 2025-06-12 docs/manuscript/paper.Rmd
Clean up old text

Before exogenous feeding, poikilothermic organisms have a near mechanistic relationship between ambient temperature and developmental rates. As such, statistical models can be easily developed to predict when organisms develop. Until recently, most models only used non-variable developmental regimes making them difficult to apply to wild environments. However, the R package hatchR formalized an approach using effective values where each day is given a developmental unit, accurately predicting developmental phenology for wild poikilotherms. hatchR was developed specific to fish, however this manuscript broadens the tool's application showing how it can be used to predict developmental phenology for a broad range of taxa, including amphibians, reptiles, and invertebrates. Moreover, we provide numerous examples of how this approach informs scenarios from applied management to basic questions regarding ecological and evolutionary questions.

BMM's version:

@@

MORE CITATIONS

Because poikilothermic organisms rely on the environment to determine their body temperature, the physiological processes that regulate their bodies is tightly linked with their ambient temperature. When these organisms are in their early life histories prior to exogenous feeding, developmental rates are solely determined by environmental and genetic and/or maternal effects, with the environmental responsible for most of the variation in phenotypes [@west-eberhard2003]. In turn, while developmental rates may vary widely across species, they typically fit a consistent non-linear relationship with temperature which can be easily mathematically approximated using the power law.

**BMM--**Because poikilothermic organisms rely on external sources to regulate their body temperature, the physiological processes governing their growth and development are tightly coupled to ambient thermal conditions. This relationship is especially deterministic during early life stages prior to exogenous feeding, when metabolic rates are primarily governed by environmental temperatures, modulated by genetic and maternal effects [@west-eberhard2003]. Although the absolute rate of development can vary widely among species and populations, most poikilotherms exhibit a consistent and often non-linear thermal performance curve that describes development as a function of temperature [@schoolfield1981; @angilletta2006]. These relationships can be effectively approximated with power law or exponential models that capture increasing developmental rates with rising temperature, up to a physiological optimum beyond which performance declines.

Historically, predicting developmental rates for populations in the wild relied on thermal thresholds where degree days are summed from mating and phenology occurs once a thermal threshold is passed. However, these models can be error prone, especially for longer developing species, because of the non-linearity of development, where embryos developing at colder temperatures need far fewer degree days to hatch than those at warm temperatures. Similarly, when applied to wild environments models fail to account for diel variabilty, where daily (and hence developmental rates) may be highly different across early life history development. Finally, populations may be locally adapted with developmental rates varying across populations [@sparks2017].

**BMM--**Historically, models of developmental phenology in wild populations have relied on the accumulation of degree days (thermal sums) above a species-specific threshold. While widely used, degree-day models are fundamentally linear and fail to account for the non-linear nature of development, leading to biased predictions under fluctuating temperature regimes [@neuheimer2009]. This is particularly problematic for long-developing species or for organisms in variable climates, where embryos incubating at colder temperatures often require fewer thermal units to complete development than those incubating in warmer environments. Moreover, traditional models often assume temporally invariant developmental rates and do not incorporate diel temperature variability, which can significantly affect development in systems with large day–night thermal swings. Finally, developmental reaction norms may vary across populations due to local adaptation or phenotypic plasticity, further complicating predictions [@sparks2017; @angilletta2009].

To address these issues in fishes, ***Sparks et al.*** (XXX)developed hatchR, which leverages effective value models developed by @sparks2019. The models function using developmental curves fit from experiments where organisms are raised at different temperatures and reciprocating those models to provide daily development estimates and are readily applied to wild environments. hatchR is packaged as an available R package on CRAN, as well as a Shiny app presented as a graphical-user-interface. While hatchR was developed specifically for fishes, the approach functions equally well for any poikilotherm for which development roughly follows a power law relationship. Here, we demonstrate the wide applicability of hatchR to non-fish organisms providing numerous putative sources for parameterizing developmental models across broad taxa and three specific examples of how hatchR can be used to investigate questions regarding ecological and inter- and intraspecies relationships.

**BMM--**To address these limitations in fishes, [@sparks2019] introduced the effective value model—an approach that captures the non-linear, temperature-dependent nature of development by integrating experimental data into a daily developmental rate function. Building on that work, the R package **hatchR** [CITE hatchR] was developed to operationalize the effective value framework and make it accessible to researchers and managers. **hatchR** uses temperature-performance curves generated from lab-based thermal rearing trials to estimate the proportion of development achieved each day under observed temperature conditions. These daily contributions are summed until a cumulative threshold (typically 1.0) is reached, marking the completion of a developmental stage. The package includes both an R-based workflow and a user-friendly Shiny web interface to facilitate broad adoption across research and management communities.

**BMM--**While hatchR was originally developed for coldwater fishes, the approach is broadly applicable to any poikilothermic organism whose development conforms to a monotonic or unimodal temperature–rate relationship, including many amphibians, reptiles, and invertebrates. In this manuscript, we expand the taxonomic scope of **hatchR**, compiling thermal developmental data from the literature and demonstrating how effective value modeling can be applied to diverse taxa. We also present three case studies illustrating how **hatchR** can inform ecological, evolutionary, and management questions—ranging from understanding thermal adaptation and phenological mismatches to forecasting hatch timing in conservation planning. These applications are increasingly relevant as freshwater ecosystems face intensifying and interacting stressors, including climate change, habitat alteration, and invasive species, all of which can disrupt developmental timing and population dynamics [@craig2017; ADD ANOTHER BROAD PAPER].

@@

As species diverge from one another it is expected that phenotypes respond to different niche space which is part of that divergence. To assess those differences,one can execyte experiments that test phenotypic divergence (common garden or reciprocal transplant) and display phenotypic data as reaction norms, showing the relative genetic and environmental contribution to phenotypes. hatchR is a unique tool for demonstrating these differences because the strong fits of developmental rates, either displayed as model fits or the linearized effective value model lines are forms of reaction norms, even if species did not experience the same developmental conditions. While not as precise as using controlled experiments, hatchR derived reaction norms can be used to demonstrate how genetic x environment interactions vary among species and how those phenotypes may fit the niche spaces occupied by phylogenetically close or divergent species.

To demonstrate this approach, we leverage custom time to hatch parameterizations for four *Lithobates* frog species developed by @moore1939. We show large differences in developmental rates for common North American frogs which often overlap in their ranges—green frog, *L. clamitans*; northern leopard frog, *L. pipiens*; pickerel frog, *L. palustris*; and wood frog, *L. sylvaticus*. Specifically, when developmental models were fit to a randomly generated temperature regime of 16 and 24 C°, the species-specific effects of devlopment are clear. Whereas at warm temperatures, all species hatch the same day or within one day of one another, at cold temperatures, wood frogs develop the quickest and green frogs the slowest (twice the rate of wood frogs).

BRIAN ADD SOMETHING ABOUT THEIR ECOLOGY ABOVE TO FIT THIS INTO BROADER CONTEXT.

**BMM-Version**

@@

Despite being designed for fishes, hatchR is a generalizeable and simple tool for predicting developmental phenology across a wide array of poikilothermic organisms. Furthermore, the tools overcomes many of the limitations of other developmental models used for these species. Tools such as hatchR will be particularly useful for interrogating species and population responses to quickening changes to their thermal environments. While we provide numerous putative sources for parameterization and a handful of case studies, we expect the applications of hatchR far extend beyond what we envision here.

**BMM-Version:**

>>>>

@@

Because poikilothermic organisms rely on external sources to regulate their body temperature, the physiological processes governing their growth and development are tightly coupled to ambient thermal conditions. This relationship is especially deterministic during early life stages prior to exogenous feeding, when metabolic rates are primarily governed by environmental temperatures, modulated by genetic and maternal effects [@west-eberhard2003]. Although the absolute rate of development can vary widely among species and populations, most poikilotherms exhibit a consistent and often non-linear thermal performance curve that describes development as a function of temperature [@schoolfield1981; @angilletta2006]. These relationships can be effectively approximated with power law or exponential models that capture increasing developmental rates with rising temperature, up to a physiological optimum beyond which performance declines.

Historically, models of developmental phenology in wild populations have relied on the accumulation of degree days (thermal sums) above a species-specific threshold. While widely used, degree-day models are fundamentally linear and fail to account for the non-linear nature of development, leading to biased predictions under fluctuating temperature regimes [@neuheimer2009]. This is particularly problematic for long-developing species or for organisms in variable climates, where embryos incubating at colder temperatures often require fewer thermal units to complete development than those incubating in warmer environments. Moreover, traditional models often assume temporally invariant developmental rates and do not incorporate diel temperature variability, which can significantly affect development in systems with large day–night thermal swings. Finally, developmental reaction norms may vary across populations due to local adaptation or phenotypic plasticity, further complicating predictions [@sparks2017; @angilletta2009].

To address these limitations in fishes, [@sparks2019] introduced the effective value model—an approach that captures the non-linear, temperature-dependent nature of development by integrating experimental data into a daily developmental rate function. Building on that work, the R package **hatchR** [CITE hatchR] was developed to operationalize the effective value framework and make it accessible to researchers and managers. **hatchR** uses temperature-performance curves generated from lab-based thermal rearing trials to estimate the proportion of development achieved each day under observed temperature conditions. These daily contributions are summed until a cumulative threshold (typically 1.0) is reached, marking the completion of a developmental stage. The package includes both an R-based workflow and a user-friendly Shiny web interface to facilitate broad adoption across research and management communities.

While hatchR was originally developed for coldwater fishes, the approach is broadly applicable to any poikilothermic organism whose development conforms to a monotonic or unimodal temperature–rate relationship, including many amphibians, reptiles, and invertebrates. In this manuscript, we expand the taxonomic scope of **hatchR**, compiling thermal developmental data from the literature and demonstrating how effective value modeling can be applied to diverse taxa. We also present three case studies illustrating how **hatchR** can inform ecological, evolutionary, and management questions—ranging from understanding thermal adaptation and phenological mismatches to forecasting hatch timing in conservation planning. These applications are increasingly relevant as freshwater ecosystems face intensifying and interacting stressors, including climate change, habitat alteration, and invasive species, all of which can disrupt developmental timing and population dynamics [@craig2017; ADD ANOTHER BROAD PAPER].

@@

@@
